Performing and evaluating graph-based clustering with rOpenScPCA

### 2024-11-14 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Continuing notebook reorg, added code for a seurat section for which glmGamPoi is now needed in renv

To begin, we'll read in the `SingleCellExperiment` (SCE) object.
We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.

```{r read data}
# Read the SCE file
sce <- readRDS(input_sce_file)

# Extract the PCA matrix
pca_matrix <- reducedDim(sce, "PCA")
```

@@

* A PCA matrix with rownames representing unique cell ids (e.g., barcodes)
* A data frame with, at least, columns representing unique cell ids and cluster assignments
  * By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters()`
  * You can override these defaults using the arguments `cell_id_col` and `cluster_col`

@@

## Working with objects directly

The above examples present how to work with a pca matrix, but you can also pass in an SCE or Seurat object with PCs to use instead.
When doing so, this package will assume a pc_name, and you will need to override this as needed.

When performing clustering on an SCE object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `PCA`.
The function assumes the PC name is `PCA`. 
If your SCE object has a different name for the principal components, you can provide the name with the argument `pc_name`.

When performing clustering on a Seurat object, `rOpenScPCA::calculate_clusters()` assumes that the principal components are stored in a reduced dimension slot named `pca`.
If your Seurat object has a different name for the principal components, you can provide the name with the argument `pc_name`.
In this example, because we created the Seurat object from the SCE object, its principal components are named `PCA`, so we will provide that information.

```{r}
# Convert to a Seurat object for demonstration
seurat_obj <- as.Seurat(
  sce,
  counts = "counts",
  data = "logcounts"
)
```

### Clustering on a Seurat object

```{r cluster seurat}
cluster_results_df <- calculate_clusters(
  seurat_obj,
  # Since we converted this from an SCE object, its PC name is `PCA`
  pc_name = "PCA"
)

cluster_results_df
```

@@

Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your scRNA-seq object.
The subsequent examples will show you, as an example, how to save the `cluster_results_df` data frame to a given object.

>>>>

To begin, we'll read in the `SingleCellExperiment` (SCE) object.

```{r read data}
# Read the SCE file
sce <- readRDS(input_sce_file)
```

We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.

```{r extract pca data}
# Extract the PCA matrix
pca_matrix <- reducedDim(sce, "PCA")
```

If you are working with a Seurat object, you can extract the PCA matrix like this:

```{r convert to seurat}
# First, convert the SCE to a Seurat object for demonstration

# Here, the PCA matrix has the same name as a
```

@@

* A PCA matrix with row names representing unique cell ids (e.g., barcodes)
* A data frame with, at least, columns representing unique cell ids and cluster assignments
  * By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters()`
  * You can override these defaults using the arguments `cell_id_col` and `cluster_col`

@@

## Evaluating pre-calculated clusters

Even if you perform clustering with other tools, you can still use the `rOpenScPCA` functions to evaluate your clusters.
For example, if you are analyzing your data with a Seurat pipeline, you might also calculate clusters with Seurat, and you can evaluate those with `rOpenScPCA`.
In addition, ScPCA data already contains a column called `cluster` with results from an automated clustering.

In this section, we'll present how you can calculate the silhouette width and neighborhood purity from existing cluster assignments.

### Evaluating Seurat clusters

First, we'll prepare a Seurat object and perform clustering for demonstration:

```{r process seurat object}
# Convert to a Seurat object
seurat_obj <- CreateSeuratObject(counts = counts(sce), assay = "RNA")

# Process the object with a Seurat pipeline to obtain clusters
seurat_obj <- seurat_obj |>
  SCTransform() |>
  RunPCA() |>
  FindNeighbors(dims = 1:30) |>
  FindClusters()

# Print resulting clusters
head(seurat_obj$seurat_clusters)
```

Next, we'll prepare inputs for `rOpenScPCA` functions:

- The PCA matrix
- A data frame with columns `cell_id` and `cluster`

```{r prepare seurat input}
# Extract the PCA matrix
seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")

# Create a data frame
seurat_cluster_df <- data.frame(
  cell_id = names(seurat_obj$seurat_clusters),
  cluster = seurat_obj$seurat_clusters
)
```

Now, we can run `calculate_silhouette()` and `calculate_purity()`:

```{r seurat silhouette}
seurat_silhouette_df <- calculate_silhouette(
  seurat_pca_matrix,
  seurat_cluster_df
)
seurat_silhouette_df
```

```{r seurat purity}
seurat_purity_df <- calculate_purity(
  seurat_pca_matrix,
  seurat_cluster_df
)
seurat_purity_df
```

### Evaluating SCE clusters

```{r}
```

* Louvain algorithm
* Jaccard weighting
* 20 nearest neighbors
* Resolution of 1

Seurat parameters:
* louvain
* resolution of 0.8
*

There are a few common circumstances where you might encounter this:

```{r}
# Print the cluster column from the SCE object
head(colData(sce)$cluster)
```

In this section, we'll demonstrate how you can run the `rOpenScPCA` cluster evaluation functions on these clusters.

First, you should prepare a data frame two columns: `cell_id` containing cell barcodes, and `cluster` containing the cluster identities.

```{r}
scpca_cluster_df <- data.frame(
  cell_id = colnames(sce),
  cluster = colData(sce)$cluster
)

scpca_cluster_df
```

Then, you can proceed to run evaluation functions using this data frame, and the PCA matrix from the object.

```{r}
# Calculate silhouette width
scpca_silhouette_df <- calculate_silhouette(pca_matrix, scpca_cluster_df)
scpca_silhouette_df

# Calculate neighborhood purity
scpca_purity_df <- calculate_purity(pca_matrix, scpca_cluster_df)
scpca_purity_df
```

When running `calculate_stability()`, it is important to specify the same parameters originally used to build the clusters:

```{r}
scpca_stability_df <- calculate_stability(
  pca_matrix,
  scpca_cluster_df,
  # provide ScPCA clustering parameters
  algorithm = "louvain",
  weighting = "jaccard",
  nn = 20,
  resolution = 1
)
scpca_stability_df
```

### Working with objects directly

As presented above, `rOpenScPCA` clustering functions take a PCA matrix with row names representing unique cell ids as their first argument.

Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains PCs.
For example:

```{r}
# Calculate clusters from an SCE object using default parameters
cluster_results_df <- calculate_clusters(sce)
```

`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects.
If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.
For example:

```{r}
# First, we'll rename the PCA matrix for demonstration
reducedDimNames(sce) <- c("PCA_matrix", "UMAP")

# Calculate clusters from an SCE object using default parameters
cluster_results_df <- calculate_clusters(
  sce,
  pc_name = "PCA_matrix"
)
```

@@

Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your SCE or Seurat object.
The subsequent examples will show you, as an example, how to save the `cluster_results_df` data frame to a given object.
